High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-01-24 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[R. Martinez Brunner et al.]{
Rafael Martinez--Brunner,$^{1}$\thanks{E-mail: rmartinezbrunner@gmail.com}
Simon Casassus,$^{1}$
Sebasti\'an P\'erez,$^{2}$
Antonio Hales,
Philipp Weber, $^{1,2}$ \newauthor
Miguel Carcamo,
Carla Arce-Tord,$^{1}$
Lucas Cieza,
Antonio Garufi,
Sebasti\'an Marino,$^{3}$
Alice Zurlo
\\
% List of institutions
$^{1}$Departamento de Astronom\'{\i}a, Universidad de Chile, Casilla 36-D, Santiago, Chile\\
$^{2}$Universidad de Santiago de Chile, Av. Ecuador 3659, Santiago\\
$^{3}$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK\\
}

@@

For the vertical structure, \citet{dOrazi} found flaring angles, i.e. $H/r$, of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values were set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is $H_\circ = 2.2$\,au at $r_\circ = 18$\,au. This is summarized in Fig.\ref{fig:profiles} bottom panel that shows the aspect ratio profile $h(r)=H(r)/r$.

@@

For Ring24 we used the same power-law as for the inner disc but scaled by an empirically obtained factor, $\delta_{\mathrm{sd}}(r)$= 4.2$\times 10^4$ for $r>18$\,au and $\delta_{\mathrm{sd}}(r)$= 0 for $r<18$\,au:
\begin{equation}
  \Sigma_{\mathrm{Ring24}}(r) = \Sigma_{\mathrm{inner\,disc}}(r)\, \delta_{\mathrm{sd}}(r),
\end{equation}
The inner edge of Ring24 follows a different profile. Between $R_\mathrm{in}$ and $R_\mathrm{peak}$, which respectively mark  the inner edge and the location of maximum density of the outer ring, the gas density profile is multiplied by an additional factor
\begin{equation}
    \epsilon(r) = \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3.
\end{equation}
This parameter allows us to model a smoother inner edge of the outer ring, with $R_\mathrm{in,gas}$ = 18.0\,au and $R_\mathrm{peak,gas}$ = 26.4\,au. This factor is also applied to the large dust population with $R_\mathrm{in,ld}$ = 24.2\,au and $R_\mathrm{peak,ld} = R_\mathrm{peak,gas}$.

@@

For Ring5 and Ring13, we choose Gaussian profiles parameterized with centroid radii $\mu_{\mathrm{Ring5, ld}}=5.5$\,au and $\mu_{\mathrm{Ring13, ld}}=13.22$\,au, ring widths of $\sigma_{\mathrm{Ring5,ld}}=0.4$\,au and $\sigma_{\mathrm{Ring13,ld}}=0.85$\,au, and normalizations $\Sigma_{\circ,\mathrm{Ring5,ld}}=1.3\times10^{-4}\,\mathrm{au\,g\,cm^{-3}}$ and $\Sigma_{\circ,\mathrm{Ring13,ld}}=2.4\,\mathrm{au\,g\,cm^{-3}}$. For Ring24 we used a similar profile as for the gas --a power-law function--, but with the difference that, in an effort to recreate the break seen in the outer ring, we choose different values for $\gamma = -3.5$ and $R_c$ = 39\,au between 27.9 and 35.3\,au. The surface density for large dust grains in the outer ring is thus given by
\begin{equation}
    \Sigma_{\mathrm{Ring24,ld}}(r) = \Sigma_{\mathrm{c}} \left(\frac{r}{R_{\mathrm{c}}(r)}\right)^{-\gamma(r)} \exp\left[-\left(\frac{r}{R_{\mathrm{c}}(r)}\right)^{2-\gamma}\right]\,\delta_{\mathrm{ld}}(r),
\end{equation}
where $\delta_{\mathrm{ld}}(r)$ ensures continuity
\begin{equation}
  \delta_{\mathrm{ld}}(r) =
  \begin{cases}
  0                 & r < 24.6\text{\,au} \\
  7.19 \times10^{4} & 24.6 < r < 27.9\text{\,au} \\
  3.17 \times10^{4} & 27.9 < r < 35.3\text{\,au} \\
  2.65 \times10^{5} & 35.3 < r < 64\text{\,au.} \\
  \end{cases}
\end{equation}
Then the final density of large dust can be calculated as
\begin{equation}
\rho_{\mathrm{large\,dust}}(r,z)=
\frac{\Sigma_{\mathrm{ld}}(r)}{\sqrt{2\pi} \, H(r)} \mathrm{exp}\left[-\frac{1}{2} \left(\frac{z}{H(r)}\right)^2\right] \, f_{\mathrm{ld}} \: \zeta .
\end{equation}

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. The introduced Ring5 is not visible at the simulated image as its peak predicted intensities are around two times the noise in the ALMA image ($\sim$1$\times10^{-7}$\,Jy\,$\mathrm{beam}^{-1}$). This gives us an upper limit for the total millimeter-sized dust mass present in Ring5 of $\sim$2$\times10^{-5}$\,M$_{\earth}$. The depletion of large dust grains in Ring5 is consistent with efficient dust trapping in Ring13. A zone of radially increasing gas pressure can entirely filter out large dust grains from inner regions, while smaller grains can be able to overcome this barrier due to their strong frictional coupling to the gas accretion flow \citep[studied in the context of planetary gaps,][]{Rice2006,Zhu2012,Weber2018}. Still, the strong depletion of large particles demands that within Ring5 dust evolution is very inefficient or limited, or that large particles are very rapidly removed. Otherwise, the present small grains should quickly coagulate to form detectable grain sizes in the ALMA observation in Fig.\ref{fig:images_vs_simulated}  \citep{Drazkowska2019}.

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. Accordingly, taking the parametric model values for Ring13, it has a radius of 13.22\,au, a FWHM of 2.00\,au, and a scale height FWHM of the millimetre-size dust of 0.61\,au. The total dust mass of Ring13 is about 0.7\,M$_{\earth}$ in the model. In turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and break at $\sim$36\,au. It contains a total dust mass of $\sim$44\,M$_{\earth}$. The model predictions for the millimetre-size dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since  the  scale height of the large grains in Ring13 is  0.61\,au, and has a  measured width of 2.46$\pm$0.56\,au, the ring is $\sim$4.1$\pm$0.9 times more extended radially than vertically.

@@

Even though the radial spread of large dust grains in Ring13 appears to be quite thin, the width in comparison to the sub-lying gas profile speaks for the presence of considerable turbulent diffusion. Following a similar ring analysis as in \citet{2018ApJ...869L..46D}, we find that the ratio between the dimensionless diffusion parameter, $D$, and the dimensionless Stokes number, St, (which parameterizes the dynamical behaviour of a grain) is roughly, $D/{\rm St}\approx 0.1$. The observed signal is expected to be dominated by grains of size $a\approx 0.02\,{\rm cm}$. The RT model, together with the dust-to-gas ratio of 0.05, prescribe a gas density of $\Sigma_{\rm g} \approx 0.5\,{\rm g}{\rm cm}^{-2}$ to the location of Ring13. With these values, the relevant Stokes number is approximated to be St$\approx 0.1$. This yields an estimate for the level of diffusivity of $D\approx 0.01$. It further provides a value for the level of turbulent viscosity in Ring13, $\alpha_{\rm turb}\approx 0.01$, assuming viscosity to be equal to the level of diffusion \citep{2007Icar..192..588Y}. We note that an observation of molecular line broadening has found no evidence for turbulent contributions, suggesting $\alpha_{\rm turb}<0.01$ \citep{Flaherty_2020}. The value inferred from our model is just within this limit. It is linearly proportional to the local dust-to-gas ratio (by our definition of the gas surface density profile) -- a standard ISM  value of 0.01 for the dust-to-gas ratio would thus lead to an expected level of turbulent viscosity of $\alpha_{\rm turb}\approx 5\times10^{-3}$. While the exact value for $\alpha_{\rm turb}$ is not well constrained, the presence of turbulence is a robust detection, as it is required to explain the radial spread of the resolved Ring13.

The observed SED is compared with the model in Fig.~\ref{fig:SED}. From the similarity with the data we propose that there has to be a small-grain population close to the stars down to 0.3\,au. The observed central and compact signal  in the ALMA observation cannot be explained using  millimeter-sized dust, following the mass estimates in Section\,\ref{sec:Observations}, as it reaches $\sim$18 times large surface densities than in the gas. Also, the hypothesis of thermal emission from large grains in the inner disc results in  a large excess at 10$\,\micron$ that does not match the data. The decision of employing the three ringed structure (i.e. add Ring5 to the observed structure) relies on the fact that the SED needed a ring at a radius around 5--7\,au to have a proper fit around 10\,$\micron$.

%Ring5 may be just artificial and not a real prediction, as a thin ring like the one we implemented would be expected to trap millimetre-size dust grains, while those grains are not present in the ALMA observation.

@@

S.C. acknowledges support from FONDECYT grant 1171624. S.P. acknowledges support from the Joint Committee of ESO and the Government of Chile and FONDECYT grant 1191934.

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[R. Martinez Brunner et al.]{Rafael Martinez--Brunner,$^{1}$
\thanks{E-mail: rmartinezbrunner@gmail.com}
Simon Casassus,$^{1}$
Sebasti\'an P\'erez,$^{2}$
Antonio Hales,$^{7,8}$
Philipp Weber, $^{1,2}$ \newauthor
Miguel Carcamo,
Carla Arce-Tord,$^{1}$
Lucas Cieza,
Antonio Garufi,
Sebasti\'an Marino,$^{3}$
Alice Zurlo$^{4,5,6}$
\\
% List of institutions
$^{1}$Departamento de Astronom\'ia, Universidad de Chile, Casilla 36-D, Santiago, Chile\\
$^{2}$Departamento de F\'isica, Universidad de Santiago de Chile, Av. Ecuador 3659, Santiago\\
$^{3}$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK\\
$^{4}$N\'ucleo de Astronom\'ia, Facultad de Ingenier\'ia y Ciencias, Universidad Diego Portales, Av. Ejercito 441, Santiago, Chile\\
$^{5}$Escuela de Ingenier\'ia Industrial, Facultad de Ingenier\'ia y Ciencias, Universidad Diego Portales, Av. Ejercito 441, Santiago, Chile \\
$^{6}$Aix Marseille Universit\'e, CNRS, LAM - Laboratoire d'Astrophysique de Marseille, UMR 7326, 13388, Marseille, France  \\
$^{7}$Joint ALMA Observatory, Avenida Alonso de C\'ordova 3107, Vitacura 7630355, Santiago, Chile \\
$^{8}$National Radio Astronomy Observatory, 520 Edgemont Road, Charlottesville, VA 22903-2475, United States of America \\
}

@@

For the vertical structure, \citet{dOrazi} found flaring angles, i.e. $H/r$, of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values was set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is $H_\circ = 2.2$\,au at $r_\circ = 18$\,au. This is summarized in Fig.\ref{fig:profiles} bottom panel that shows the aspect ratio profile $h(r)=H(r)/r$.

@@

For Ring24 we used the same power-law as for the inner disc but scaled by an empirically obtained factor, $\delta_{\mathrm{sd}}(r)$:
\begin{equation}
  \Sigma_{\mathrm{Ring24}}(r) = \Sigma_{\mathrm{inner\,disc}}(r)\, \delta_{\mathrm{sd}}(r),
\end{equation}
with
\begin{equation}
  \delta_{\mathrm{sd}}(r) =
  \begin{cases}
  0    & r < 18\text{\,au} \\
  4.2\times 10^4 & r > 18\text{\,au}.
  \end{cases}
\end{equation}
The inner edge of Ring24 follows a different profile. Between $R_\mathrm{in}$ and $R_\mathrm{peak}$, which respectively mark  the inner edge and the location of maximum density of the outer ring, the gas density profile is multiplied by an additional factor
\begin{equation}
    \epsilon(r) = \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3.
\end{equation}
This parameter allows us to model a smoother inner edge of the outer ring, with $R_\mathrm{in,gas}$ = 18.0\,au and $R_\mathrm{peak,gas}$ = 26.4\,au. This factor is also applied to the large dust population with $R_\mathrm{in,ld}$ = 24.2\,au and $R_\mathrm{peak,ld} = R_\mathrm{peak,gas}$.

@@

For Ring5 and Ring13, we choose Gaussian profiles parameterized with centroid radii $\mu_{\mathrm{Ring5, ld}}=5.5$\,au and $\mu_{\mathrm{Ring13, ld}}=13.22$\,au, ring widths of $\sigma_{\mathrm{Ring5,ld}}=0.4$\,au and $\sigma_{\mathrm{Ring13,ld}}=0.85$\,au, and normalizations $\Sigma_{\circ,\mathrm{Ring5,ld}}=1.3\times10^{-4}\,\mathrm{au\,g\,cm^{-3}}$ and $\Sigma_{\circ,\mathrm{Ring13,ld}}=2.4\,\mathrm{au\,g\,cm^{-3}}$. For Ring24 we used a similar profile as for the gas --a power-law function-- but with the difference that, in an effort to recreate the break seen in the outer ring, we choose different values for $\gamma = -3.5$ and $R_c$ = 39\,au between 27.9 and 35.3\,au. The surface density for large dust grains in the outer ring is thus given by
\begin{equation}
    \Sigma_{\mathrm{Ring24,ld}}(r) = \Sigma_{\mathrm{c}} \left(\frac{r}{R_{\mathrm{c}}(r)}\right)^{-\gamma(r)} \exp\left[-\left(\frac{r}{R_{\mathrm{c}}(r)}\right)^{2-\gamma}\right]\,\delta_{\mathrm{ld}}(r),
\end{equation}
where
\begin{equation}
  \delta_{\mathrm{ld}}(r) =
  \begin{cases}
  0                 & r < 24.6\text{\,au} \\
  7.2 \times10^{4} & 24.6 < r < 27.9\text{\,au} \\
  3.2 \times10^{4} & 27.9 < r < 35.3\text{\,au} \\
  2.7 \times10^{5} & 35.3 < r < 64\text{\,au.} \\
  \end{cases}
\end{equation}
Then the final density of large dust can be calculated as
\begin{equation}
\rho_{\mathrm{large\,dust}}(r,z)=
\frac{\Sigma_{\mathrm{ld}}(r)}{\sqrt{2\pi} \, H(r)} \mathrm{exp}\left[-\frac{1}{2} \left(\frac{z}{H(r)}\right)^2\right] \, f_{\mathrm{ld}} \: \zeta .
\end{equation}

@@

The simulated 1.3\,mm continuum image reproduces the two observed rings: the faint Ring13 and the brighter Ring24. The introduced Ring5 is not visible at the simulated image as its peak predicted intensities are around two times the noise in the ALMA image ($\sim$1$\times10^{-7}$\,Jy\,$\mathrm{beam}^{-1}$). This gives us an upper limit for the total millimeter-sized dust mass present in Ring5 of $\sim$2$\times10^{-5}$\,M$_{\earth}$. The depletion of large dust grains in Ring5 is consistent with efficient dust trapping in Ring13. A zone of radially increasing gas pressure can entirely filter out large dust grains from inner regions, while smaller grains can be able to overcome this barrier due to their strong frictional coupling to the gas accretion flow \citep[studied in the context of planetary gaps,][]{Rice2006,Zhu2012,Weber2018}. Still, the strong depletion of large particles demands that within Ring5 dust evolution is very inefficient or limited, or that large particles are very rapidly removed. Otherwise, the present small grains should quickly coagulate to form detectable grain sizes in the ALMA observation in Fig.\ref{fig:images_vs_simulated}  \citep{Drazkowska2019}. The central emission is not present in the 1.3\,mm continuum image as it was impossible to achieve a bright blob without disrupting the SED fit (read below).

As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig.~\ref{fig:radprofiles}), we can assume that the model provides a possible approximation of the disc structure, including the dimensions of Ring13. Accordingly, taking the parametric model values for Ring13, it has a radius of 13.22\,au, a FWHM of 2.00\,au, and a scale height FWHM of the millimetre-size dust of 0.61\,au. The total dust mass of Ring13 is about 0.7\,M$_{\earth}$ in the model. In turn, our model reproduces the observations of Ring24 with peak intensity at $\sim$30\,au, and break at $\sim$36\,au. It contains a total dust mass of $\sim$44\,M$_{\earth}$. The model predictions for the millimetre-size dust population in Ring13 are close to the measurements, with only a 0.07\,au difference in the centroid location of the ring, and a 0.46\,au difference in the width estimation. Since  the  scale height of the large grains in Ring13 is  0.61\,au, and has a  measured width of 2.46$\pm$0.56\,au, the ring is 4.1$\pm$0.9 times more extended radially than vertically.

@@

Even though the radial spread of large dust grains in Ring13 appears to be quite thin, the width in comparison to the sub-lying gas profile speaks for the presence of considerable turbulent diffusion. Following a similar ring analysis as in \citet{2018ApJ...869L..46D}, we find that the ratio between the dimensionless diffusion parameter, $D$, and the dimensionless Stokes number, St, (which parameterizes the dynamical behaviour of a grain) is roughly, $D/{\rm St}\approx 0.1$. The observed signal is expected to be dominated by grains of size $a\approx 0.02\,{\rm cm}$. The RT model, together with the dust-to-gas ratio of 0.05, prescribe a gas density of $\Sigma_{\rm g} \approx 0.5\,{\rm g}{\rm cm}^{-2}$ to the location of Ring13. With these values, the relevant Stokes number is approximated to be St$\approx 0.1$. This yields an estimate for the level of diffusivity of $D\approx 0.01$. It further provides a value for the level of turbulent viscosity in Ring13, $\alpha_{\rm turb}\approx 0.01$, assuming viscosity to be equal to the level of diffusion \citep{2007Icar..192..588Y}. We note that an observation of molecular line broadening has found no evidence for turbulent contributions, suggesting $\alpha_{\rm turb}<0.01$ \citep{Flaherty_2020}. The value inferred from our model is just within this limit. It is linearly proportional to the local dust-to-gas ratio (by our definition of the gas surface density profile) a standard ISM  value of 0.01 for the dust-to-gas ratio would thus lead to an expected level of turbulent viscosity of $\alpha_{\rm turb}\approx 5\times10^{-3}$. While the exact value for $\alpha_{\rm turb}$ is not well constrained, the presence of turbulence is a robust detection, as it is required to explain the radial spread of the resolved Ring13.

The observed SED is compared with the model in Fig.~\ref{fig:SED}. From the similarity with the data we propose that there has to be a small-grain population close to the stars down to 0.3\,au. The observed central and compact signal in the ALMA observation cannot be explained with the hypothesis of thermal emission from large grains in the inner disc. Following the mass estimates in Section\,\ref{sec:Observations}, the millimeter-sized dust needed reaches $\sim$18 times larger surface densities than in the gas, resulting in a large excess at 10$\,\micron$ that does not match the data. On the other hand, the decision of employing the three ringed structure (i.e. add Ring5 to the observed structure) relies on the fact that the SED needed a ring at a radius around 5--7\,au to have a proper fit around 10\,$\micron$.

@@

S.C. acknowledges support from FONDECYT grant 1171624.

S.P. acknowledges support from the Joint Committee of ESO and the Government of Chile and FONDECYT grant 1191934.

A.Z. acknowledges support from the FONDECYT Iniciaci\'on en investigaci\'on project number 11190837
